\appendix

\section{Synthetic Validation Details}
\label{app:synth}

The synthetic test (Table~\ref{tab:synth}) runs with no model and no network and reproduces on
any machine, so it isolates the estimator and decision rule from any property of a particular
language model. We construct paired activations whose contrast subspace is an analytically
known object, then check four properties.

\paragraph{Construction.} For each turn we draw class-conditional activation clouds whose
difference-of-means defines a rank-$k$ ground-truth subspace with a prescribed orientation.
Two ground-truth trajectories are used: a \emph{static} subspace whose orientation is fixed
across turns, and a \emph{rotating} subspace whose orientation advances by a fixed
$15^\circ$ per turn. Estimation noise is controlled by the number of probe pairs, matching the
power regime used on real data (\S\ref{sec:res-power}).

\paragraph{What it confirms.} (i) At high SNR the estimator recovers the injected subspace.
(ii) The static trajectory returns measured drift at the bootstrap noise floor, and the
decision rule returns negative, the property that prevents a low-power run from being read as a
positive. (iii) The $15^\circ$/turn rotation is recovered as noise-corrected chordal drift
$0.144$ against the analytic $0.149$, clears the floor, and the rule
returns positive; a cross-subspace overlap with a known time-varying relative angle is tracked
over the same trajectory. (iv) Principal-angle sanity: identical planes return $0$, orthogonal
planes return $\pi/2$. Because the identical code path is used on real data, a verdict on a real
model reflects the data rather than a code artifact.

\section{Evaluation Matrix: Done and Pending}
\label{app:matrix}

Table~\ref{tab:matrix} records the evaluation matrix. The acceptance risk for a scoped negative
is not novelty but breadth, the objection that the effect was missed because of the wrong model,
space, or rank; we therefore enumerate the cells that close that objection and mark each as done
or pending, with the released harness producing the statistic for both. The single most
important pending cell is the safety subspace in SAE feature space on an instruction-tuned chat
model (e.g.\ Gemma~2 with Gemma Scope, \citealp{lieberum2024gemmascope}). It is supported by the
released harness but is pending larger compute: a \texttt{transformer\_lens} plus Gemma-2 load
issue prevented it on the available T4, and it is expected to complete on an A100-class GPU. We
mark it pending rather than report it.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}llc@{}}
\toprule
Cell & Objection it closes & Status \\
\midrule
Qwen2.5-0.5B-Instruct, raw acts, rank-1 & baseline & done \\
Qwen2.5-1.5B-Instruct, raw acts, rank-1 & ``only the smallest model'' & done \\
GPT-2 small, SAE feature space, rank-1 & ``raw acts aren't where methods operate'' & done \\
static-projector demonstration (Qwen2.5-0.5B) & ``you never tested the actual claim'' & done \\
safety subspace in SAE space, instruction-tuned (Gemma-2 + Gemma Scope) & SAE safety side on a chat model & pending (A100) \\
mixture-of-experts model (per-expert premise) & per-expert variant of the premise & pending \\
rank-$2$/$3$ subspaces & ``rank-1 is too crude'' & pending (needs hundreds of probes) \\
within-topic content control & ``topic conflated with content'' & pending \\
behavioral (generation-time) static-projector test & generation-time mis-specification & pending (future work) \\
\bottomrule
\end{tabular}
\caption{Evaluation matrix. Raw-activation replication across two model scales and the SAE
feature-space replication (GPT-2) are complete, as is the static-projector demonstration. The
instruction-tuned SAE cell for the safety subspace is supported by the released harness but
pending a larger GPU. The protocol ports to every pending cell with only a backend swap.}
\label{tab:matrix}
\end{table}

\section{Probe-Bank Construction}
\label{app:probes}

\paragraph{Contrastive pairs.} Subspaces are recovered by contrast, following standard practice
in the probing and steering literature. The \emph{hallucination/epistemic} subspace $H_t$ is the
rank-1 difference-of-means direction between answerable real-entity questions and surface-matched
unanswerable fictional-entity questions, following the knowledge-awareness construction of
\citet{ferrando2025entity}. The \emph{safety} subspace $S_t$ is the analogous contrast between
harmful requests and surface-matched harmless requests, the refusal direction of
\citet{arditi2024refusal}. Pairs are surface-matched so that the recovered direction isolates the
epistemic or safety contrast rather than lexical or syntactic differences.

\paragraph{Fixed probes, varying context.} At dialogue turn $t$ we prepend the running
conversation (the first $t$ user turns, each followed by a fixed neutral assistant reply) and
append the contrastive probe bank as the next user turn, reading the last-token residual
activation at each layer. The probes are held fixed and only the prepended context varies, so any
movement in the recovered $U_t$ is attributable to context.

\paragraph{Power-driven sizing.} The power analysis (\S\ref{sec:res-power}) requires $\geq 96$
probe pairs per side at rank-1; we use $40$--$96$ pairs per side in the reported runs. The
deliberately small banks used for laptop development are not large enough for a publishable
verdict, which is precisely why the floor is reported with every drift.

\paragraph{Scrambled-twin null.} The length- and topic-matched null is generated by
word-shuffling each user turn. This preserves the exact token multiset (and hence length, topic,
and lexicon) while destroying coherent meaning, so genuine content-conditionality requires the
real subspace to move more than its scrambled twin. As noted in the scope discussion, the
scrambled twin holds topic fixed but conflates coherent dialogue content with epistemic state; a
fully within-topic manipulation (a pending cell) would separate the two.

\section{Reproducibility Statement}
\label{app:repro}

All code, probe banks, dialogue scripts, cached activations, and the synthetic ground-truth
validation are released in a public repository. Every figure and number in this paper, including
the synthetic validation (Table~\ref{tab:synth}), the power curve (Table~\ref{tab:power}), the
rigor cascade (Table~\ref{tab:cascade}), the controlled result and its cross-model and
SAE-feature-space replications (Tables~\ref{tab:qwen05}--\ref{tab:gpt2sae}), the static-projector
demonstration (Table~\ref{tab:static}), and Figures~\ref{fig:scaled}, \ref{fig:placebo}, and
\ref{fig:static}, is regenerated by the released scripts.

\paragraph{Pipeline.} The estimator and decision rule are validated with a model-free synthetic
test that reproduces anywhere. Activation extraction reads the last-token residual stream at each
layer; the SAE-feature-space backend passes that activation through the frozen released SAE
encoder and runs the identical contrast-and-drift pipeline on the feature vector, so the only
change between the raw and SAE settings is a backend swap. Drift is computed as the
gauge-invariant chordal distance with dialogue-level bootstrap confidence intervals, and the
pre-registered decision rule (real CI lower bound $>$ scrambled CI upper bound over a
$\geq 3$-contiguous-layer band) is applied uniformly across all settings.

\paragraph{Determinism and scope.} Random seeds for the bootstrap and for synthetic data
generation are fixed, so the reported intervals reproduce. The released harness packages the
gauge-invariant drift metric, the scrambled-twin null, dialogue-level bootstrap CIs, the power
curve, and the synthetic validation as a drop-in test with a single backend swap between raw
activations and SAE features, so that a representational-drift claim, positive or negative, can be
vetted before it is built upon. All results are scoped to the models and feature spaces evaluated
here, \texttt{Qwen2.5-0.5B-Instruct} and \texttt{Qwen2.5-1.5B-Instruct} in raw activation space and
GPT-2 small in SAE feature space; we do not claim the verdict generalizes beyond them.