\documentclass{article}
\usepackage{tmlr}
% \usepackage[accepted]{tmlr}   % uncomment for the camera-ready
% \usepackage[preprint]{tmlr}   % uncomment for a non-anonymous preprint

\usepackage{amsmath,amssymb,booktabs,graphicx,xcolor}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{fillbetween}
\graphicspath{{../results/}}

\title{Is the Safety and Hallucination Feature Geometry of \\ Language Models Conditional on Dialogue Context? \\ A Confound-Controlled Study}

\author{\name Anonymous \email anon@example.com \\ \addr Affiliation withheld for review}

\def\month{06}
\def\year{2026}
\def\openreview{\url{https://openreview.net/forum?id=XXXXXXXXXX}}

\begin{document}
\maketitle

\begin{abstract}
A growing class of inference-time alignment methods, conditional activation
steering and conditional subspace disentanglement of hallucination from safety,
rests on an implicit premise: that the low-dimensional residual-stream directions
encoding safety/refusal and epistemic uncertainty are \emph{conditional} on dialogue
context, so that a single static intervention learned on one context is
mis-specified for another. We ask whether this premise is empirically true, and find
that the obvious ways of measuring it are confounded. We introduce a confound-controlled
protocol: subspaces are recovered by contrastive difference-of-means, drift is measured
as a gauge-invariant chordal distance between subspaces, and ``the subspace moved with
context'' is tested against (i) a turn-1 baseline that removes the generic
no-context$\to$context jump, (ii) a length- and topic-matched \emph{scrambled-twin}
null that holds token count and lexical content fixed while destroying coherent
meaning, and (iii) a dialogue-level bootstrap confidence interval so that no single
dialogue can carry the result. Applied to an instruction-tuned model, three intuitive
positive measurements (cumulative displacement, between-dialogue spread versus a
same-context floor, and a small-sample placebo comparison) are each overturned by the
next control. Under the full protocol, the apparent context-conditionality of the
hallucination subspace is statistically indistinguishable from a length/topic artifact:
scrambled twins move the subspace as much as real dialogues, and the real-versus-placebo
confidence intervals overlap at every layer. We further fit the static disentangling
projector these methods would replace and show its across-turn re-entanglement is fully
matched by the length/topic placebo, i.e.\ the static baseline does not genuinely fail. Beyond the specific null, the transferable
contributions are a reusable protocol, a taxonomy of three confounds that each
independently produce a false positive (a no-context onset, a same-context null that is
rigged to fire, and small-sample single-dialogue effects), and a control checklist for
anyone making a ``context changes the representation'' claim; we also derive an actionable
requirement for conditional-alignment methods, namely that they report this controlled
measurement before assuming the premise. We report the null as scoped to the models and
conditions tested. \textbf{All numbers in this draft are from
real runs on the released code. The null replicates across raw activation space (Qwen2.5-0.5B
and 1.5B) and sparse-autoencoder feature space (GPT-2 small, and Gemma-2-2B-it for \emph{both}
the hallucination and the safety/refusal subspace), the representation disentanglement methods
operate in, where the scrambled twin matches or exceeds the real spread at several layers. Only
the mixture-of-experts cell remains in progress.}
\end{abstract}

\section{Introduction}
\label{sec:intro}

Mechanistic interpretability has established that high-level behaviors of large language
models (LLMs) are often approximately linearly encoded: refusal is mediated by a single
residual-stream direction \citep{arditi2024refusal}, truth/falsehood is linearly
represented \citep{marks2023geometry}, and sparse autoencoders (SAEs) recover
interpretable feature directions at scale \citep{bricken2023monosemanticity,
templeton2024scaling, lieberum2024gemmascope}. These directions can be read and steered
\citep{zou2023repe, li2023iti, panickssery2024caa} and erased in closed form
\citep{ravfogel2020inlp, belrose2023leace}.

A natural next step, pursued by recent work, is to \emph{disentangle} behaviors that
share latent components: for example, to mitigate hallucination without degrading safety
by orthogonalizing the SAE subspaces that encode them \citep{mahmoud2025unintended}. Once
one moves from a single static edit to a \emph{conditional} one, an implicit empirical
premise enters: that the relevant geometry is conditional on context, so that a single
projection learned on one context (or on single-turn data) is mis-specified elsewhere,
and in particular drifts as a multi-turn dialogue accumulates. This premise is rarely
stated, let alone tested. It is also non-obvious: the geometry-of-refusal literature has
shown that geometric orthogonality need not imply behavioral independence
\citep{wollschlager2025geometry}, that refusal is multi-directional \citep{pan2025hidden},
and that steering vectors generalize poorly and are brittle \citep{tan2024steering}.

We ask the prior question directly: \emph{does dialogue context actually rotate the
safety and hallucination subspaces, beyond trivial confounds?} The contribution is not a
new alignment method but a measurement and a caution:

\begin{itemize}
\item \textbf{A confound-controlled protocol} (\S\ref{sec:method}) for testing
representational drift: gauge-invariant chordal drift between contrastively-recovered
subspaces, a turn-1 baseline, a length/topic-matched scrambled-twin null, and
dialogue-level bootstrap confidence intervals. The protocol is validated against analytic
ground truth.
\item \textbf{A rigor cascade} (\S\ref{sec:cascade}) showing that three intuitive
positive measurements of context-conditionality are each artifacts: a no-context onset
jump, a same-context null that carries no length/topic variance, and a small-sample
($n{=}3$ dialogue) placebo comparison.
\item \textbf{A robustness finding} (\S\ref{sec:main}): under the full protocol the
hallucination subspace's apparent context-conditionality is statistically indistinguishable
from a length/topic artifact, and this null \emph{replicates} across two raw model scales and in
SAE feature space for \emph{both} the hallucination and the safety/refusal subspace (the latter
on Gemma-2-2B-it), the representation conditional disentanglement actually uses.
\item \textbf{Actionable recommendations} (\S\ref{sec:lessons}): a control checklist for
representational-drift claims, and the requirement that conditional-alignment methods report
the controlled premise measurement (and a failing static baseline) before introducing
conditional machinery.
\end{itemize}

We frame this as a robustness/methodology study rather than a single negative number: the
generalizable output is the protocol, the confound taxonomy, and the recommendations, with
the replicated null as the evidence that they matter.

Our claim is deliberately scoped: we do not claim the premise is false for all models or
in SAE feature space. We claim that the premise is \emph{not supported} by the obvious
measurements once they are controlled, that the controls matter (they flip the
conclusion), and that methods built on this premise should report the controlled
measurement before assuming it.

\input{related.tex}

\input{method.tex}

\input{results.tex}

\section{Discussion, Scope, and Limitations}
\label{sec:limits}

\paragraph{What this does and does not establish.} It establishes that, for the model and
construction tested, three intuitive measurements of context-conditionality are
confounded, and that the controlled measurement returns a null. It does \emph{not}
establish that the premise is universally false. We have closed the most important cells and
mark the remainder. The instruction-tuned SAE-feature-space cell is now run on Gemma-2-2B-it for
\emph{both} the hallucination and the safety/refusal subspace (Section~\ref{sec:main},
Table~\ref{tab:gemma}), the only setting in which the safety subspace is recoverable enough to
test, and it too returns the null; and the static-projector mis-specification test
(Section~\ref{sec:res-static}) shows the static baseline does not genuinely fail. The genuinely
open cells are: \textbf{(a)} larger instruction-tuned models and a mixture-of-experts model (for
the per-expert variant of the premise); \textbf{(b)} higher-rank subspaces (rank-1 is the powered
default); \textbf{(c)} a within-topic content manipulation that separates epistemic state from
surface topic; and \textbf{(d)} a behavioral, generation-time static-projector test. A caveat on
the Gemma cell: the public Gemma Scope SAEs are base-model-trained, so the instruction-tuned run
is a mild distribution mismatch and is indicative. Our protocol ports to all of these with only a
change of backend.

\paragraph{Residual confounds in the control.} The scrambled twin holds token length and
topic but conflates coherent ``dialogue content'' with ``epistemic state''; a fully
within-topic content manipulation would separate these. Word-shuffling may also partially
degrade topic extraction; this would, if anything, bias toward a (spurious) positive,
which we do not observe.

\paragraph{Implication.} The negative result is not an indictment of conditional steering
or disentanglement; it is a request that such methods measure, with controls, the premise
they assume. A static intervention is mis-specified only if the geometry it freezes
actually moves with context beyond length and topic. On the present evidence, the cheap
and decisive thing is to run the controlled measurement first.

\section{Lessons and Actionable Recommendations}
\label{sec:lessons}

The point of this study is not the single null but the transferable lessons that produced
it; we state them as recommendations for two audiences.

\paragraph{For anyone measuring whether a manipulation changes a representation}
(context, in-context examples, a steering edit, a fine-tune), the following four controls
flip conclusions in practice and should be standard:
\begin{enumerate}
\item \textbf{Use a matched-null, not a same-condition resampling floor.} A bootstrap over
probes at a fixed condition contains none of the between-condition variance (length, position,
topic) that the test statistic carries, so ``effect $>$ floor'' is near-tautological. Build a
null that matches the nuisance variables and differs only in the variable of interest (here,
length/topic-matched scrambled twins). This single change converted our headline from
positive to null.
\item \textbf{Measure with a gauge-invariant quantity.} Subspaces are recovered up to an
arbitrary basis; a basis-dependent ``drift'' conflates estimator gauge with real movement.
Use principal-angle/chordal distances.
\item \textbf{Bootstrap over the unit of generalization, not the convenient one.} A
confidence interval over probes says nothing about generalization across dialogues; resample
the dialogues. Our mid-layer ``effect'' was significant over probes and vanished over
dialogues ($n{=}3 \to n{=}12$).
\item \textbf{Baseline away trivial transitions.} The no-context$\to$context jump is large,
shared, and uninteresting; measure drift relative to the first in-context turn, and report the
onset separately.
\end{enumerate}

\paragraph{For designers of conditional alignment methods} (conditional steering, conditional
disentanglement): a conditional intervention is justified only if the geometry it conditions on
\emph{moves} with context beyond nuisance confounds, and if a static baseline \emph{fails}. We
recommend reporting, as a precondition, (i) the controlled context-conditionality measurement
above, in the feature space the method uses, and (ii) a static-projector baseline tested for
content-driven re-entanglement. We carried out exactly this test (\S\ref{sec:main},
Figure~\ref{fig:static}): the static projector's across-turn leak is fully matched by the
length/topic placebo, so it does not genuinely fail. On the present evidence the conditional
machinery, for the behaviors and models studied, addresses a confound rather than a real
non-stationarity. We do not claim this generalizes to all settings; we claim that without this
controlled check a method cannot tell the two apart, and the check is cheap.

\paragraph{A reusable checklist.} The released harness packages the above as a drop-in test:
a gauge-invariant drift metric, a scrambled-twin null, dialogue-level bootstrap CIs, a power
curve, and a synthetic ground-truth validation, with a single backend swap between raw
activations and SAE features. The intended use is pre-registration: run it, report the
controlled measurement and the power, and only then build (or not).

\section{Conclusion}

We asked whether dialogue context rotates the safety and hallucination subspaces of an LLM,
the premise underlying conditional disentanglement. The obvious measurements say yes; each
is a confound. Under a length/topic-matched, dialogue-level-bootstrapped protocol the
effect is within noise on the model tested, and the safety subspace is not recoverable
where the premise would need to hold. We release the protocol and the full rigor cascade so
that representational-drift claims, positive or negative, can be vetted before they are
built upon.

\paragraph{Reproducibility.} All code, probe banks, dialogues, cached activations, and the
synthetic ground-truth validation are released; every figure and number is regenerated by
the released scripts.

\input{appendix.tex}

\bibliography{refs}
\bibliographystyle{tmlr}

\end{document}
